EGGs are adhesive!

### 2025-03-04 calco/main.tex
minima

E-graphs (shortly, EGGs) are an up-and-coming formalism for program optimisation and synthesis via a compact 
representation and efficient implementation of equality saturation. 
%
Albeit a classical data structure~\cite{DetlefsNS05}, EGGs received 
new impulse after the seminal~\cite{WillseyNWFTP21} and
developed a thriving community, as witnessed by the official website~\cite{eggs}.
%
The key idea of rewriting-based program optimisation is to perform the manipulation of a syntactical description 
of a program, replacing some of its components in such a way that the semantics is preserved while improving 
e.g. time or space costs of its actual computation. Instead of directly removing sub-programs, EGGs just add the 
new components and link them to the older ones via the equivalence relation, until an optimal program is 
reached and extracted.

@@

The aim of this work is to describe EGGs by means of adhesive categories. The advantages are two-fold. On the one side, 
we put the benefits \emph{per se} of a formal presentation, making precise the properties of the data structure. 
On the other side, describing the addition of equivalent sub-programs 
in terms of DPO transformation offer the possibility of modelling parallel execution of optimisation steps as well as discussing
about confluence and termination.

\emph{Synopsis}
The paper has the following structure. 
In Section~\ref{sec:ade} we briefly recall 
the theory of $\mathcal{M}$-adhesive categories
and of kernel pairs.
In Section~\ref{sec:hyper} we present the graphical structures of our interest, 
 (labelled, acyclic) hypergraphs and term graphs, and we provide a
functorial characterisation, which allows for proving their adhesivity properties.
This is expanded in Section~\ref{hypereq} for describing hypergraphs 
and term graphs with equivalence and in Section~\ref{eggs} for capturing
their variants where the equivalences are closed with respect to operator application,
thus subsuming EGGs.
%
In Section~\ref{rewriting} we put the former machinery at work, showing how the optimisation steps
can be rephrased as suitable applications of term graph rewriting rules.
%
Finally, in Section~\ref{conclusioni} we draw our conclusions, hint at future endeavours and offer some 
brief remarks on related works.
%
For the sake of space, the proofs appear in the appendices. 
%For the sake of completeness, 
%and in order to fix the notation, we prove all the results 
%recalled in the background section, besides those that are original 
%to our work.

@@

What about rewriting? The theory of $\mathcal{M},\mathcal{N}$-adhesivity ensures that if the rules are spans of 
arrows in $\mathcal{M}$ and the match is an arrow in $\mathcal{N}$, then we can lift all the standard properties
of the DPO approach to the category at hand. Luckily, this is the structure of the rules that is advanced in~\cite{DetlefsNS05},
which is given as a span $L \leftarrow L \rightarrow R$, where the first component is the identity and the second
component is a regular mono, simply adding the edges, nodes, and equivalence classes of interest.
%
An EGG rule such $x \div x \to 1$, from the introductory example in~\cite{WillseyNWFTP21}, 
can be modelled as the DPO rule below,
which can be applied with a match that is injective on equivalence classes to the left-most and the middle EGG
in the picture above
\[\mbox{to draw DPO rule, x in the same equivalence class} \]

@@

Summing up, showing that EGGs are objects of $\GEqTGs$ and that optimisation steps are obtained via 
DPO rules with negative application conditions unleash the xxx.

>>>>

E-graphs (shortly, EGGs) are an up-and-coming formalism for program optimisation and synthesis via a compact 
representation and efficient implementation of equality saturation. 
%
Albeit a classical data structure~\cite{DetlefsNS05}, EGGs received 
new impulse after the seminal~\cite{WillseyNWFTP21} and
developed a thriving community, as witnessed by the official website~\cite{eggs}.
%
The key idea of rewriting-based program optimisation is to perform the manipulation of a syntactical description 
of a program, replacing some of its components in such a way that the semantics is preserved while improving 
the computational costs of its actual execution. Instead of directly removing sub-programs, EGGs just add the 
new components and link them to the older ones via the equivalence relation, until an optimal program is 
reached and extracted.

@@

The aim of this work is to prove that EGGs form an adhesive category. The advantages are two-fold. On the one side, 
we put the benefits \emph{per se} of a formal presentation, making precise the properties of the data structure. 
On the other side, describing the addition of equivalent sub-programs 
in terms of DPO transformation offer the possibility of modelling parallelism and concurrency for optimisation steps 
and discussing confluence and termination.

\emph{Synopsis}
The paper has the following structure. 
In Section~\ref{sec:ade} we briefly recall 
the theory of $\mathcal{M}$-adhesive categories
and of kernel pairs.
In Section~\ref{sec:hyper} we present the graphical structures of our interest, 
 (labelled) hypergraphs and term graphs, and we provide a
functorial characterisation, which allows for proving their adhesivity properties.
This is expanded in Section~\ref{hypereq} for describing hypergraphs 
and term graphs with equivalence and in Section~\ref{eggs} for capturing
their variants where the equivalences are closed with respect to operator application,
thus subsuming EGGs.
%
In Section~\ref{rewriting} we put the former machinery at work, showing how the optimisation steps
can be rephrased as suitable applications of term graph rewriting rules.
%
Finally, in Section~\ref{conclusioni} we draw our conclusions, hint at future endeavours and offer some 
brief remarks on related works.
%
For the sake of space, the proofs appear in the appendices. 
%For the sake of completeness, 
%and in order to fix the notation, we prove all the results 
%recalled in the background section, besides those that are original 
%to our work.

@@

What about rewriting? The theory of $\mathcal{M},\mathcal{N}$-adhesivity ensures that if the rules are spans of 
arrows in $\mathcal{M}$ and the match is an arrow in $\mathcal{N}$, then we can lift all the standard properties
of the DPO approach to the category at hand. Luckily, this is the structure of rules that is advanced in~\cite{DetlefsNS05},
which is given as a span $L \leftarrow L \rightarrow R$, where the first component is the identity and the second
component is a regular mono, simply adding the edges, nodes, and equivalence classes of interest.
%
An EGG rule such $x \div x \to 1$, from the introductory example in~\cite{WillseyNWFTP21}, 
can be modelled as the DPO rule below,
which can be applied with a match that is injective on equivalence classes to the left-most and the middle EGG
in the picture above
\[\mbox{to draw DPO rule, x in the same equivalence class} \]

@@

Summing up, showing that EGGs are objects of an $\mathcal{M},\mathcal{N}$-adhesive category $\GEqTGs$ and that optimisation 
steps are obtained via DPO rules with negative application conditions with rules whose arrows are in $\mathcal{M}$ and via a match 
that is in $\mathcal{N}$ allows for
